---
title: "Google_Divvy_Analytics"
author: "Bijaya"
date: "2023-10-17"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This is a data analysis with the data sets from a Chicago based bike rental company Cyclistic. It owns approx 5800 bikes and 692 docking stations across the city.I being a junior data analyst, have analysed the data and derived insights based on the points put forward by the stakeholders.

Data source: `https://divvy-tripdata.s3.amazonaws.com/index.html`  
- Downloaded files: 
  - Divvy_Trips_2019_Q1.zip
  - Divvy_Trips_2019_Q2.zip 
  - Divvy_Trips_2019_Q3.zip
  - Divvy_Trips_2019_Q4.zip

```{r}
options(repos = c(CRAN = "https://cran.r-project.org"))
```

```{r}
install.packages("RMySQL")
install.packages("DBI")
install.packages("dbplyr")
```

```{r}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
library(dplyr)
library(RMySQL)
library(DBI)
getwd() #displays your working directory
```
### loading data
```{r}
# Upload Divvy datasets (csv files) here
q1 <- read_csv("Divvy_Trips_2019_Q1.csv", show_col_types = FALSE)
q2 <- read_csv("Divvy_Trips_2019_Q2.csv", show_col_types = FALSE)
q3 <- read_csv("Divvy_Trips_2019_Q3.csv", show_col_types = FALSE)
q4 <- read_csv("Divvy_Trips_2019_Q4.csv", show_col_types = FALSE)
```
### All dataframes combined
```{r}
all_trips <- bind_rows(q1, q2, q3, q4)
```

### Summary
```{r}
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```
# Empty cells
```{r}
colSums(is.na(all_trips)) #Check for missing cells in data frame
```
### Remove unwanted cols first
### then rows with empty cells
```{r}
all_trips <- subset(all_trips, select = - c(trip_id, bike_id, gender, birth_year)) 
all_trips <- na.omit(all_trips)
colnames(all_trips)
```

### check to confirm rows with empty cells are removed

```{r}
all_trips[!complete.cases(all_trips), ]
```
# Rename Colnames for convenience 
```{r}
all_trips <- rename(all_trips
                    ,started_at = start_time 
                    ,ended_at = end_time
                    ,start_station_name = from_station_name 
                    ,start_station_id = from_station_id
                    ,end_station_name = to_station_name 
                    ,end_station_id = to_station_id 
                    ,member_casual = user_type
                    )
```
### convert date-time column types into `POSIXct`
```{r}
all_trips$started_at <- as.POSIXct(all_trips$started_at, format = "%m/%d/%y %H:%M")
all_trips$ended_at <- as.POSIXct(all_trips$ended_at, format = "%m/%d/%y %H:%M")
```
### remove rows with "`ended_at` < `started_at`" values
```{r}
all_trips <- subset(all_trips, started_at < ended_at)
dim(all_trips)
```
### check if above conversion took place
```{r}
str(all_trips$started_at)
str(all_trips$ended_at)
```
### add a column for day of the week
```{r}
all_trips$month_name <- months(as.Date(all_trips$started_at))
all_trips$day <- format(as.Date(all_trips$started_at), "%d")
all_trips$year <- format(as.Date(all_trips$started_at), "%Y")
all_trips$day_name <- format(as.Date(all_trips$started_at), "%A")
```
### check for how `day_name` column will look like
```{r}
random_sample <- all_trips[sample(nrow(all_trips), 10), ]
random_sample
```
### Minimum
```{r}
min(all_trips$trip_duration_sec)
```
### Reassign to the desired column to easy-to-remember values
```{r}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
```
### Table by user_type
```{r}
yearly_count_by_user <- all_trips %>% 
  group_by(member_casual) %>%
    summarise(count = n())
yearly_count_by_user 
```
### Bar Plot for User Count in 2019 
```{r}
# Create a ggplot for the df_mean data frame
ggplot(yearly_count_by_user, aes(x = member_casual, y = count, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(
    title = "User Count by Type",
    x = "Member Type",
    y = "User Count"
  ) +
  theme_minimal()
```
### Pie Chart Representation
```{r}
yearly_count_by_user$percent = paste(round((yearly_count_by_user $count /sum(yearly_count_by_user $count))*100), "%")

ggplot(yearly_count_by_user , aes(x = "", y = percent, fill = member_casual)) +
  geom_col(color = "black") +
  geom_text(aes(label = percent),
            position = position_stack(vjust = 0.5)) +
  coord_polar(theta = "y") +
  theme_void()  + labs(title = "Yearly User Type Percentage")
```


### Mean ride_duration by user_type
```{r}
yearly_mean_sec_by_user <- all_trips %>%
  group_by(member_casual) %>%
    summarise(mean_duration_sec = round(mean(trip_duration_sec)))
yearly_mean_sec_by_user
```
### Bar Plot yearly mean trip duration by user_type
```{r}
# Create a ggplot for the df_mean data frame
ggplot(yearly_mean_sec_by_user, aes(x = member_casual, y = mean_duration_sec, fill = member_casual)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Mean Trip Duration by Member Type",
    x = "Member Type",
    y = "Mean Duration (seconds)"
  ) +
  theme_minimal()
```

### Bar chart for `rental count vs day of the week`
```{r}
day_of_week_total <- all_trips %>%
  group_by(day_name) %>%
    summarise(count = n())
day_of_week_total
```
### Bar Plot average trip count against day of week 
```{r}
# Specify the order of days of the week
day_order <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

# Convert day_name to a factor with the desired order
day_of_week_total$day_name <- factor(day_of_week_total$day_name, levels = day_order)

ggplot(day_of_week_total, aes(x = day_name, y = count, fill = "day_name")) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "2019 Rentals by Day and Member Type", x = "Day Name", y = "Count") +
  theme_minimal()
```
### day of the week by Member Type
```{r}
day_of_week_by_user <- all_trips %>%
  group_by(day_name, member_casual) %>%
  summarise(count = n(), .groups = "drop")
day_of_week_by_user
```
### Bar Plot average trip count for day of week by user type
```{r}
# Create the bar chart
# Convert day_name to a factor with the desired order
day_of_week_by_user$day_name <- factor(day_of_week_by_user$day_name, levels = day_order)

ggplot(day_of_week_by_user, aes(x = day_name, y = count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "2019 Rentals by Day and Member Type", x = "Day Name", y = "Count") +
  theme_minimal()
```
### total trips each month
```{r}
month_total <- all_trips %>%
  group_by(month_name) %>%
    summarise(count = n())
month_total
```
### Bar Plot total trips each month
```{r}
# Specify the order of days of the week
month_order <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Convert day_name to a factor with the desired order
month_total$month_name <- factor(month_total$month_name, levels = month_order)

ggplot(month_total, aes(x = month_name, y = count, fill = month_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Bar Chart of Rentals by Month and Member Type", x = "Month Name", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust the angle as needed
```
### Each month total by user type
```{r}
month_by_user <- all_trips %>%
  group_by(month_name, member_casual) %>%
    summarise(count = n())
month_by_user$month_name <- factor(month_by_user$month_name, levels = month_order)
month_by_user
```
### Bar Plot Each month trip count by user type
```{r}
# Convert day_name to a factor with the desired order
month_by_user$month_name <- factor(month_by_user$month_name, levels = month_order)

ggplot(month_by_user, aes(x = month_name, y = count, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "2019 Rentals by Month and Member Type", x = "Month Name", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust the angle as needed
```

### Data migration to MySQL database
### DataFrame `all_trips` is migrated to MySQL table `Divvy_Trips_2019` under database `Divvy_Case_Study

```{r}
con <- dbConnect(
  RMySQL::MySQL(),
  dbname = "Divvy_Case_Study",
  host = "localhost",
  username = "root",
  password = "Mnandhar2015_mac"    # encrypted for privacy
)

dbWriteTable(con, name = 'yearly_count_by_user', value = yearly_count_by_user, overwrite = TRUE)
dbWriteTable(con, name = 'yearly_mean_sec_by_user', value = yearly_mean_sec_by_user, overwrite = TRUE)
dbWriteTable(con, name = 'day_of_week_total', value = day_of_week_total, overwrite = TRUE)
dbWriteTable(con, name = 'day_of_week_by_user', value = day_of_week_by_user, overwrite = TRUE)
dbWriteTable(con, name = 'month_total', value = month_total, overwrite = TRUE)
dbWriteTable(con, name = 'month_by_user', value = month_by_user, overwrite = TRUE)
dbWriteTable(con, name = 'Divvy_Trips_2019', value = all_trips, overwrite = TRUE)


dbDisconnect(con)
```